---
title: "Cluster identification - Assembloid epithelial cells"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---


```{r, message=FALSE}
rm(list=ls())
library(Seurat)
library(ggplot2)

library(readxl)
library(writexl)
library(knitr)
library(data.table)
library(cowplot)
library(magrittr)
library(dplyr)

library(pheatmap)

library(grid)

data_storage_folder = ("../../Results/epithelial/")
result_folder = data_storage_folder

library(future)
options("future.globals.maxSize"=2**33)
plan("multisession", workers = 8)

rerun_dge_analyses = T

```



```{r}
scData.combined = readRDS(file.path(data_storage_folder, "Assembloids_Organoids_Tissue_combined_SCT_Harmony_epithelial.rds"))

set.seed(1234)
# scData.combined <- RunUMAP(scData.combined, reduction = "harmony", dims = 1:25, verbose = FALSE, assay="SCT")
# scData.combined <- FindNeighbors(scData.combined, reduction = "harmony", dims = 1:25, assay="SCT")
# scData.combined <- FindClusters(scData.combined, resolution = 0.2)  # modify resolution to change the number of clusters

```


# Cluster overview

```{r, fig.width=10, fig.height=6}
DimPlot(scData.combined, reduction = "umap", label = TRUE)
```

```{r, fig.width=10, fig.height=6}
DimPlot(scData.combined, reduction = "umap", label = TRUE, split.by = "Condition", ncol = 4)
```


```{r}
m = table(scData.combined$seurat_clusters, scData.combined$Condition)
m
```


```{r}
prop.table(m, margin=2) |> pheatmap()
```


# Final cluster assigment

```{r}
cluster_assignment = read.table("../../Data/cluster_assignments/epithelial/Assembloid_tissue_organoid_cluster_SCT_Harmony_2023-02-14.txt", sep="\t",header=T, stringsAsFactors = F)
options(width = 180)
print(cluster_assignment)
rownames(cluster_assignment) = as.character(cluster_assignment$ClusterID)
scData.combined = AddMetaData(scData.combined, cluster_assignment[as.character(scData.combined$seurat_clusters),"Label"], col.name = "cluster_label")
scData.combined = AddMetaData(scData.combined, cluster_assignment[as.character(scData.combined$seurat_clusters),"Type"], col.name = "cluster_type")
```

```{r, fig.width=14, fig.height=8}
scData.combined = SetIdent(scData.combined, value="cluster_label" )
p = DimPlot(scData.combined, reduction = "umap", label = TRUE)
print(p)
```

```{r, fig.width=8, fig.height=4}
p = DimPlot(scData.combined, reduction = "umap", label = TRUE , split.by = "Condition") + NoLegend()
print(p)
```


## Distribution of clusters within samples


```{r}
m = table(scData.combined$cluster_label, scData.combined$SampleName)
m
```

```{r, fig.width=10,fig.height=7}
mm  = as.matrix.data.frame(m)
colnames(mm) = colnames(m)
rownames(mm) = rownames(m)

mm_r = sweep(mm, 2, apply(mm,2,sum), "/")

pheatmap(mm_r, cluster_cols = F, cluster_rows=F, main="Proportion of each cluster per sample")
```


## Distribution of clusters within conditions

```{r}
m = table(scData.combined$cluster_label, scData.combined$Condition)
m
```

```{r}
mm_r = prop.table(mm, margin = 2)
options(width=200)
kable(round(mm_r,3))
```


```{r, fig.width=10,fig.height=7}
pheatmap(mm_r, cluster_cols = F, cluster_rows=F)
```

```{r, fig.width=10, fig.height=8}
mm_rel_ts = reshape2::melt(t(mm_r))
colnames(mm_rel_ts) = c("Group","Cluster","Proportion_cells")
mm_rel_ts$Cluster = factor(paste0("",mm_rel_ts$Cluster))

#mm_rel_ts$Tissue = ifelse(grepl("BM|Knochenmark", mm_rel_ts$Group), "BM","Liver")

ggplot(mm_rel_ts, aes(x=Group, y = Proportion_cells, group=Cluster)) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle=60, hjust=1)) + facet_wrap(~ Cluster )

```


```{r}
annotated_output_file = file.path(data_storage_folder, "Assembloids_organoids_tissue_annotated_SCT_Harmony_annotated.RData")
if(!file.exists(annotated_output_file)) save(scData.combined,file=annotated_output_file, compress=T)
```


# Export

Cluster assignations have been exported to file *Metadata_all_combined_SCT_Harmony.txt*.

```{r}
tmp = scData.combined@meta.data
tmp$cell_id = rownames(tmp)
write.table( tmp , file=file.path(data_storage_folder,"Metadata_all_combined_SCT_Harmony.txt"), sep="\t", row.names = F, quote = F)
```




# Marker genes for clusters

```{r}
if(rerun_dge_analyses) {
  cluster_markers <- FindAllMarkers(object = scData.combined, only.pos = TRUE, assay = "RNA")
  save(cluster_markers, file=file.path(data_storage_folder, "All_cluster_markers_SCT_Harmony.Rdata"))
  write_xlsx(cluster_markers,file.path(result_folder, "All_cluster_markers_SCT_Harmony.xlsx"))
} else {
  load(file.path(data_storage_folder, "All_cluster_markers_SCT_Harmony.Rdata"))
}
```

```{r}
plot_top_genes <- function(data, dge_results_all, title) {
  for (n in sort(unique(dge_results_all$cluster))) {
    tmp = subset(dge_results_all, cluster==n)
    if(nrow(tmp)<1) next
    tmp_up = subset(tmp, avg_log2FC > 0)
    if(nrow(tmp_up)<1) next
    tmp_up$direction = "up"

    final_top_genes = head(tmp_up, 10)
    final_top_genes = final_top_genes[order(final_top_genes$avg_log2FC),]
    final_top_genes$gene = factor(final_top_genes$gene, levels = final_top_genes$gene)
    p = ggplot(final_top_genes, aes(y=avg_log2FC, x=gene, fill=direction)) + geom_bar(stat="identity") + ggtitle(paste0(n, ", ", title)) + coord_flip() + theme(axis.text.y =  element_text(size=12) )
    print(p)
    
  }
}
```



## Percent expression of top markers across clusters

```{r, fig.width=16, fig.height=14}
top_5_markers = unique(unlist(tapply(cluster_markers$gene, cluster_markers$cluster, head, 5)))
DotPlot(scData.combined, features = top_5_markers , dot.scale = 4, cols = c("red", "blue"), assay = "RNA") +    RotatedAxis()
```

## Log2 Fold Changes of top 10 DE genes from each cluster

```{r, fig.width=12, fig.height=6}
plot_top_genes(scData.combined, cluster_markers, "Cluster markers")
```


```{r}

plot_genes <- function(dataset, genes) {
  
  all_genes =rownames(dataset@assays$RNA)
  
  for (g in genes) {
    if (!g %in% all_genes) {
      print(paste0("Gene ",g," has no data (possibly filtered out)."))
      next
    } else {
        p1  =FeaturePlot(object = dataset, features = g, reduction="umap", label = T)
        p2 = VlnPlot(dataset, features = g, ncol = 1, pt.size = 0.01)
        print(plot_grid(p1,p2, ncol=2,align="h"))
    }
  }
}
```


# Individual marker expression

## Global markers

These are global cell markers for major cell lineages and types (B/T-cells, Mo/MF etc)

Cells were sorted as CD14+ cells.

CD45 is Ptprc.  
Pglyrp1 - circulating polymorphonuclear leukocytes (granulocytes + mast cells)  

We here add Cd3e/d/g as pan-T marker and Cd20 (Ms4a1) as mature B-cell marker.
Gata3, Cxcr6, Rora - T-cell markers

Ly6a(Sca-1) - positive in HSC/MPP
Procr,Fdg5, Hoxb5 - HSC

Epor, Slamf1(Cd150) - MegE - lineage

```{r, fig.width=14, fig.height=6}

sel_markers=c("Ptprc","Pglyrp1","Cd4","Cd3e","Cd3d","Cd3g", "Gata3","Cxcr6","Rora","Cd7","Cd8a","Cd19","Ms4a1","Ly6a", "Cd34","Procr","Fgd5","Hoxb5","Epor","Slamf1","Epcam","Vim","Col1a1","Pecam1","Vwf")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)

```

```{r, fig.width=14, fig.height=3}
DotPlot(scData.combined, features = sel_markers, cols="RdYlBu", assay = "RNA")
subset(scData.combined, Condition == "tissue") |> DotPlot(features = sel_markers, cols="RdYlBu", assay = "RNA") + ggtitle("Tissue only")
```
## Colon epithelial cell markers

```{r, fig.width=14, fig.height=6}
sel_markers=c("Lgr5","Ascl2","Axin2","Aqp8","Muc2","Spink4","Dclk1","Tff3","Chga","Krt20")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
```

```{r, fig.width=12, fig.height=3}
DotPlot(scData.combined, features = sel_markers, cols="RdYlBu", assay = "RNA") + ggtitle("All conditions")
subset(scData.combined, Condition == "tissue") |> DotPlot(features = sel_markers, cols="RdYlBu", assay = "RNA") + ggtitle("Tissue only")
subset(scData.combined, Condition == "assembloids") |> DotPlot(features = sel_markers, cols="RdYlBu", assay = "RNA") + ggtitle("Assembloids only")
subset(scData.combined, Condition == "organoids") |> DotPlot(features = sel_markers, cols="RdYlBu", assay = "RNA") + ggtitle("Organoids only")
```

## Proliferation

```{r, fig.width=14, fig.height=6}
sel_markers=c("Top2a","Mki67","Pcna")
DefaultAssay(scData.combined) <- "RNA"

#plot_genes(scData.combined, sel_markers)
```

```{r, fig.width=12, fig.height=3}
DotPlot(scData.combined, features = sel_markers, cols="RdYlBu", assay = "RNA")
```


## Mitochondrial transcript proportion

```{r}
FeaturePlot(object = scData.combined,  features = "percent_mito", reduction="umap", label=T)
```

```{r, fig.height=8}
tmp = data.frame(percent_mito = scData.combined$percent_mito, batch= scData.combined$orig.ident, group_label = scData.combined$SampleName )
ggplot(tmp, aes(fill=batch, x=group_label, y = percent_mito)) + geom_boxplot() + theme(axis.text.x = element_text(angle=90, hjust=1))

```


## Scores and RNA count per cluster

```{r, fig.width=10, fig.height=4}
VlnPlot(scData.combined, features = "percent_mito", ncol = 1, pt.size = 0.01)
#VlnPlot(scData.combined, features = "HALLMARK_APOPTOSIS", ncol = 1, pt.size = 0.01)
VlnPlot(scData.combined, features = "nCount_RNA", ncol = 1, pt.size = 0.01)
VlnPlot(scData.combined, features = "nFeature_RNA", ncol = 1, pt.size = 0.01)
VlnPlot(scData.combined, features = "nFeature_RNA", ncol = 1, pt.size = 0.01, split.by="Condition")
```

```{r, fig.width=8, fig.height=4}
FeaturePlot(object = scData.combined,  features = "nFeature_RNA", reduction="umap", label=T)
FeaturePlot(object = scData.combined,  features = "nFeature_RNA", reduction="umap", label=T, split.by = "Condition")
```



# Software versions

```{r}
sessionInfo()
```
