---
title: "Assembloids epithelial cells - Data import"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---

```{r, message=FALSE}
rm(list=ls())
library(Seurat)
library(ggplot2)

library(readxl)
library(knitr)
library(data.table)
library(cowplot)


data_storage_folder = file.path("../../Results/epithelial")
if (!file.exists(data_storage_folder)) dir.create(data_storage_folder, recursive=T)
```

# Overview

scRNA seq count data from assembloid epithelial cells, measured on a 10x.  


## Sample description

```{r}
ed = as.data.frame(read_excel("../../Data/sample_descriptions/sample_names_epithelial.xlsx", sheet=1))

ed

rownames(ed) = ed$SampleID
```


```{r}
  mat_files = list.files(path="../../Data/filtered_matrices/epithelial", pattern = "*.mtx.gz", recursive = T, full.names = T)
  folders = dirname(mat_files)
  sample_ids = unlist(sapply(strsplit(folders,"/"), function(x) x[length(x)] ) )
  names(folders) = sample_ids
  
  all_sample_data = list()
  for (n in sample_ids) {
    dir(path=folders[n])
    tmp = Read10X( folders[n] )
    all_sample_data[[n]] = CreateSeuratObject(tmp, project=n)
  }
  
  all_ge_data_raw = merge(all_sample_data[[1]], all_sample_data[2:length(all_sample_data)])
  rm(all_sample_data)
  gc(verbose=F)
  
  all_ge_data_raw$cell_id = Cells(all_ge_data_raw)
  all_ge_data_raw$orig.ident = strsplit(all_ge_data_raw$cell_id, "_") |> sapply(function(x) paste(x[1:2], collapse="_") ) |> unlist()
  all_ge_data_raw$sampleID = all_ge_data_raw$orig.ident
  all_ge_data_raw$SampleName = ed[all_ge_data_raw$orig.ident, "SampleName"]
  all_ge_data_raw$Condition = ed[all_ge_data_raw$orig.ident, "Condition"]
  all_ge_data_raw$Replicate = ed[all_ge_data_raw$orig.ident, "Replicate"]

```


## Cells per sample (unfiltered)

```{r}
meta = all_ge_data_raw@meta.data
table(meta$orig.ident)
barplot(table(meta$orig.ident), las=2, ylab="Cell number")
```

# Quality control

## Detected genes

```{r, fig.width = 8}
ggplot(meta) + geom_boxplot(aes(x=orig.ident, y=nFeature_RNA, fill=SampleName)) + theme(axis.text.x = element_text(angle=45, hjust=1)) + ylab("Number of detected genes") + xlab("Sample") + scale_y_log10()
```

## UMIs

```{r, fig.width = 8}
ggplot(meta) + geom_boxplot(aes(x=orig.ident, y=nCount_RNA, fill=SampleName)) + theme(axis.text.x = element_text(angle=45, hjust=1)) + ylab("Total RNA counts") + xlab("Sample") + scale_y_log10()
```


## Overview

```{r}
all_ge_data_raw <- PercentageFeatureSet(all_ge_data_raw, "^MT-|^mt-", col.name = "percent_mito")
all_ge_data_raw <- PercentageFeatureSet(all_ge_data_raw, "^RP[SL]|^Rp[sl]", col.name = "percent_ribo")
```

### General features

```{r, fig.width=14, fig.height=10}
feats <- c("nFeature_RNA","nCount_RNA","percent_mito","percent_ribo")
VlnPlot(all_ge_data_raw, group.by= "SampleName", features = feats, pt.size = 0,ncol = 2) + NoLegend()
```

### Pairwise parameter correlations

```{r, fig.width=14, fig.height=12}
cowplot::plot_grid(ncol = 2, nrow = 2, 
  FeatureScatter(all_ge_data_raw, "nCount_RNA"  , "nFeature_RNA", group.by = "SampleName", pt.size = .5) + guides(colour = guide_legend(override.aes = list(size=4))),
  FeatureScatter(all_ge_data_raw, "percent_mito", "nFeature_RNA", group.by = "SampleName", pt.size = .5) + guides(colour = guide_legend(override.aes = list(size=4))),
  FeatureScatter(all_ge_data_raw, "percent_ribo", "nFeature_RNA", group.by = "SampleName", pt.size = .5) + guides(colour = guide_legend(override.aes = list(size=4))),
  FeatureScatter(all_ge_data_raw, "percent_ribo", "percent_mito", group.by = "SampleName", pt.size = .5) + guides(colour = guide_legend(override.aes = list(size=4)))
)

```

### Pairwise feature distribution densities

```{r, fig.width=14, fig.height=10}
meta = as.data.table(all_ge_data_raw@meta.data)

ggplot(meta) + geom_density2d_filled(aes(x=nCount_RNA, y=nFeature_RNA), contour_var = "ndensity" ) + facet_wrap(~SampleName) + ggtitle("nCount_RNA")
ggplot(meta) + geom_density2d_filled(aes(x=percent_ribo, y=nFeature_RNA), contour_var = "ndensity" ) + facet_wrap(~SampleName) + ggtitle("percent ribo")
ggplot(meta) + geom_density2d_filled(aes(x=percent_mito, y=nFeature_RNA), contour_var = "ndensity" ) + facet_wrap(~SampleName) + ggtitle("percent mitochondrial")
```

### Thresholds for numbers of detected genes and percentage of mitochondrial genes

```{r}
hist(meta$nFeature_RNA, 1000, main="Number of detected genes per cell"); abline(v=c(300,4000), col="red")
hist(meta$percent_mito, 1000, main="Percent mitochondrial RNA"); abline(v=5, col="red")
```


```{r}
per_sample_min_counts = c("SP087_019"=2000,
                          "SP087_020"=1500,
                          "SP087_021"=1500,
                          "SP087_022"=1500,
                          "SP087_023"=1500,
                          "SP087_024"=1500)
```


```{r}
ggplot(meta, aes(x=nFeature_RNA)) + geom_density() + facet_wrap(~SampleName) + geom_vline(xintercept=c(unique(per_sample_min_counts),9000), lty=2) + ggtitle("Number of detected features")
ggplot(meta, aes(x=nCount_RNA)) + geom_density() + facet_wrap(~SampleName) + geom_vline(xintercept=c(3000), lty=2) + ggtitle("Total UMIs") + scale_x_log10()
ggplot(meta, aes(x=percent_mito)) + geom_density() + facet_wrap(~SampleName) + geom_vline(xintercept=c(20), lty=2) + ggtitle("Percent mitochondrial transcripts")
```


# Filtering 

We exclude cells with less than a sample specific number of detected genes (see below), those with > 9000 detected genes (possible doublets) and those with clear signs of cellular stress (percentage of mitochondrial RNA > 20%). 

```{r}
per_sample_min_counts

#feature_low = 1500
feature_high = 9000
max_mito = 20

all_ge_data_raw = AddMetaData(all_ge_data_raw, (all_ge_data_raw$nFeature_RNA > per_sample_min_counts[all_ge_data_raw$orig.ident]) & (all_ge_data_raw$nFeature_RNA < feature_high) & (all_ge_data_raw$percent_mito < max_mito), col.name="CellFilterKeep")

ribo.genes <- grepl(pattern = "^RP[SL]|^Rp[sl]", x = rownames(x = all_ge_data_raw@assays$RNA))
mito.genes <- grepl(pattern = "^MT-|^mt-", x = rownames(x = all_ge_data_raw@assays$RNA))

selected_c <- WhichCells(all_ge_data_raw, expression = CellFilterKeep )
selected_f <- rownames(all_ge_data_raw)[ Matrix::rowSums(all_ge_data_raw) >= 2 & !(ribo.genes | mito.genes)]
```


## Number of removed/kept cells per condition

```{r}
m = table(all_ge_data_raw$SampleName, ifelse(all_ge_data_raw$CellFilterKeep, "Keep","Remove") )
m
```

```{r}
barplot(t(m), col=c("blue","red"), main="Filtering results")
legend("topright", legend=c("Kept","Removed"), fill = c("blue","red"))
```


```{r}
all_ge_data_filtered <- subset(all_ge_data_raw, features=selected_f, cells=selected_c)

dim(all_ge_data_filtered)
```


```{r, fig.width=14, fig.height=10}
ggplot(all_ge_data_filtered@meta.data) + geom_density2d_filled(aes(x=percent_mito, y=log10(nCount_RNA)), contour_var = "ndensity" ) + facet_wrap(~SampleName) + ggtitle("percent mitochondrial")
```


Raw and filtered data is saved to individual files *All_Filtered_scData.Rdata*.

```{r}
save(all_ge_data_filtered, ed, file=file.path(data_storage_folder, "All_Filtered_scData.Rdata"))
```


# Software versions

```{r}
sessionInfo()
```

