---
title: "GSE114374 colon mesenchyme - Data import, mouse"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---

```{r, message=FALSE}
rm(list=ls())
library(Seurat)
library(ggplot2)

library(readxl)
library(knitr)
library(data.table)
library(cowplot)

#options(connectionObserver = NULL) # temporary fix for RStudio issue with RSQLite
#library(org.Hs.eg.db)
library(rtracklayer)

data_storage_folder = "../../../Results/Kinchen_Stroma"

if (!file.exists(data_storage_folder)) dir.create(data_storage_folder, recursive=T)

options("save.defaults" = list(compress=TRUE))

```

# Overview

Data is from GSE114374 (mouse colon mesenchyme scRNA sequencing). Cells were processed using 10x flow cells.  


# Data import

```{r}
  all_input_files = list.files("../../../Data/external/GSE114374_Kinchen_Stroma", pattern = "*txt.gz", full.names = T)
  tmp = strsplit(gsub("_expression_matrix\\.txt\\.gz", "", basename(all_input_files)),"_")
  
  ed =  as.data.frame(t(as.data.frame(tmp)))
  colnames(ed) = c("GSE","Species","Condition")
  ed$file = all_input_files
  ed$ID = paste0(ed$Species, "_", ed$Condition)
  rownames(ed) <- ed$ID
  
  sel_data = subset(ed, Species == "Mouse")
  
  all_sample_data = list()
  
  for (fid in sel_data$ID) {
    data_tab = fread(cmd = paste0("gzip -dc ",ed[fid, "file"]), sep="\t")
    data_mat = as.sparse(data_tab[, 2:ncol(data_tab), with=F])
    rownames(data_mat) = data_tab$V1
    rm(data_tab)
    
    all_sample_data[[fid]] = CreateSeuratObject(counts = data_mat, project = fid)
    
  }
  
  all_ge_data_raw = merge(all_sample_data[[1]], all_sample_data[2:length(all_sample_data)])
  rm(all_sample_data)
  gc(verbose=F)

  all_ge_data_raw$group = all_ge_data_raw$orig.ident
  all_ge_data_raw$orig.ident = unlist(sapply(strsplit(Cells(all_ge_data_raw), "-"), `[`,2))
  all_ge_data_raw$condition = unlist(sapply(strsplit(all_ge_data_raw$group, "_"), `[`,2))
  
  all_ge_data_raw$sample_label = paste0(all_ge_data_raw$condition, "_", all_ge_data_raw$orig.ident)


```

## Cells per sample (unfiltered)

```{r}
meta = all_ge_data_raw@meta.data
with(meta, table(orig.ident , dnn="Sample"))
with(meta, table(condition , dnn="Sample type"))

```

# Quality control

## Detected genes

```{r, fig.width = 8}
ggplot(meta) + geom_boxplot(aes(x=orig.ident, y=nFeature_RNA, fill=condition)) + theme(axis.text.x = element_text(angle=45, hjust=1)) + ylab("Number of detected genes") + xlab("Sample") + scale_y_log10()
```

## UMIs

```{r, fig.width = 8}
ggplot(meta) + geom_boxplot(aes(x=orig.ident, y=nCount_RNA, fill=condition)) + theme(axis.text.x = element_text(angle=45, hjust=1)) + ylab("Total RNA counts") + xlab("Sample") + scale_y_log10()
```


## Overview

```{r}
all_ge_data_raw <- PercentageFeatureSet(all_ge_data_raw, "^MT-|^mt-", col.name = "percent_mito")
all_ge_data_raw <- PercentageFeatureSet(all_ge_data_raw, "^RP[SL]|^Rp[sl]", col.name = "percent_ribo")
meta = all_ge_data_raw@meta.data
```

### General features

```{r, fig.width=14, fig.height=10}
feats <- c("nFeature_RNA","nCount_RNA","percent_mito","percent_ribo")
VlnPlot(all_ge_data_raw, group.by= "condition", features = feats, pt.size = 0.01,ncol = 2) + NoLegend()
```

### Pairwise parameter correlations

```{r, fig.width=14, fig.height=12}
cowplot::plot_grid(ncol = 2, nrow = 2, 
  FeatureScatter(all_ge_data_raw, "nCount_RNA"  , "nFeature_RNA", group.by = "condition", pt.size = .5) + guides(colour = guide_legend(override.aes = list(size=4))),
  FeatureScatter(all_ge_data_raw, "percent_mito", "nFeature_RNA", group.by = "condition", pt.size = .5) + guides(colour = guide_legend(override.aes = list(size=4))),
  FeatureScatter(all_ge_data_raw, "percent_ribo", "nFeature_RNA", group.by = "condition", pt.size = .5) + guides(colour = guide_legend(override.aes = list(size=4))),
  FeatureScatter(all_ge_data_raw, "percent_ribo", "percent_mito", group.by = "condition", pt.size = .5) + guides(colour = guide_legend(override.aes = list(size=4)))
)

```

<!-- ### Pairwise feature distribution densities -->

<!-- ```{r, fig.width=14, fig.height=10} -->
<!-- meta = as.data.table(all_ge_data_raw@meta.data) -->

<!-- ggplot(meta) + geom_density2d_filled(aes(x=nCount_RNA, y=nFeature_RNA), contour_var = "ndensity" ) + facet_wrap(~Condition) + ggtitle("nCount_RNA") -->
<!-- ggplot(meta) + geom_density2d_filled(aes(x=percent_ribo, y=nFeature_RNA), contour_var = "ndensity" ) + facet_wrap(~Condition) + ggtitle("percent ribo") -->
<!-- #ggplot(meta) + geom_density2d_filled(aes(x=percent_mito, y=nFeature_RNA), contour_var = "ndensity" ) + facet_wrap(~group) + ggtitle("percent mitochondrial") -->
<!-- ``` -->

### Thresholds for numbers of detected genes and percentage of mitochondrial genes

```{r}
hist(meta$nFeature_RNA, 1000, main="Number of detected genes per cell"); abline(v=c(500,5000), col="red")
hist(meta$percent_mito, 1000, main="Percent mitochondrial RNA"); abline(v=5, col="red")
```

```{r}
ggplot(meta, aes(x=percent_mito))  + geom_density() + facet_wrap(~sample_label) + geom_vline(xintercept = 5,  col="red")
```


# Filtering 

We exclude cells with less than 500 detected genes, those with > 2500 detected genes (possible doublets) in CD_I/II and > 3500 detected genes in BM_I (accounting for possible higher expressed gene counts in progenitors) and those with clear signs of cellular stress (percentage of mitochondrial RNA > 15%). 

```{r}
feature_low = 500
feature_high = 5000
max_mito = 5

all_ge_data_raw = AddMetaData(all_ge_data_raw, (all_ge_data_raw$nFeature_RNA > feature_low) & (all_ge_data_raw$nFeature_RNA < feature_high) & (all_ge_data_raw$percent_mito < max_mito), col.name="CellFilterKeep")

ribo.genes <- grepl(pattern = "^RP[SL]|^Rp[sl]", x = rownames(x = all_ge_data_raw@assays$RNA))
mito.genes <- grepl(pattern = "^MT-|^mt-", x = rownames(x = all_ge_data_raw@assays$RNA))

selected_c <- WhichCells(all_ge_data_raw, expression = CellFilterKeep )
selected_f <- rownames(all_ge_data_raw)[ Matrix::rowSums(all_ge_data_raw) >= 2 & !(ribo.genes | mito.genes)]
```


## Number of removed/kept cells per Condition

```{r}
table(all_ge_data_raw$condition, ifelse(all_ge_data_raw$CellFilterKeep, "Keep","Remove") )
table(all_ge_data_raw$sample_label, ifelse(all_ge_data_raw$CellFilterKeep, "Keep","Remove") )
```


```{r}
all_ge_data_filtered <- subset(all_ge_data_raw, features=selected_f, cells=selected_c)

dim(all_ge_data_filtered)
```

Raw and filtered data is saved to individual files *All_Filtered_scData.Rdata*.

```{r}
save(all_ge_data_filtered, ed, file=file.path(data_storage_folder, "All_Filtered_scData.Rdata"), compress = T)
```


# Software versions

```{r}
sessionInfo()
```

