---
title: "Cluster identification - GSE172261 Colon mesenchyme, mouse"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---


```{r, message=FALSE}
rm(list=ls())
library(Seurat)
library(ggplot2)

library(readxl)
library(writexl)
library(knitr)
library(data.table)
library(cowplot)
library(magrittr)
library(dplyr)

library(homologene)

library(pheatmap)

library(grid)
library(png)

data_storage_folder = "../../../Results/Jasso_Stroma"
result_folder = data_storage_folder

library(future)
options("future.globals.maxSize"=2**33)
plan("multisession", workers = 8)

options("save.defaults" = list(compress=TRUE))


rerun_analyses = T
```



```{r}
load(file.path(data_storage_folder, "GSE172261_ColonMesenchyme_mouse.combined.RData"))

set.seed(1234)

```


# Cluster overview

```{r, fig.width=14, fig.height=8}
DimPlot(scData.combined, reduction = "umap", label = TRUE)
```

```{r, fig.width=14, fig.height=12}
DimPlot(scData.combined, reduction = "umap", label = TRUE, split.by = "condition", ncol = 4)
```

# Marker genes for clusters

```{r}
if(rerun_analyses) {
  cluster_markers <- FindAllMarkers(object = scData.combined, only.pos = TRUE, assay = "RNA")
  save(cluster_markers, file=file.path(data_storage_folder, "All_markers.Rdata"))
  write_xlsx(cluster_markers,file.path(result_folder, "All_markers.xlsx"))
} else {
  load(file.path(data_storage_folder, "All_markers.Rdata"))
}
```

```{r}
plot_top_genes <- function(data, dge_results_all, title) {
  for (n in sort(unique(data$seurat_clusters))) {
    tmp = subset(dge_results_all, cluster==n)
    if(nrow(tmp)<1) next
    tmp_up = subset(tmp, avg_log2FC > 0)
    if(nrow(tmp_up)<1) next
    tmp_up$direction = "up"
    #tmp_down = subset(tmp, avg_log2FC < 0)
    #tmp_down$direction = "down"
    
    #final_top_genes = rbind(head(tmp_up,10), head(tmp_down,10))
    final_top_genes = head(tmp_up, 10)
    final_top_genes = final_top_genes[order(final_top_genes$avg_log2FC),]
    final_top_genes$gene = factor(final_top_genes$gene, levels = final_top_genes$gene)
    p = ggplot(final_top_genes, aes(y=avg_log2FC, x=gene, fill=direction)) + geom_bar(stat="identity") + ggtitle(paste0(n, ", ", title)) + coord_flip() + theme(axis.text.y =  element_text(size=12) )
    print(p)
    
  }
}
```


```{r}
cluster_markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> top10
```


## overview (10 top genes for each cluster)

```{r, fig.width=14, fig.height=12}
DoHeatmap(scData.combined, features = top10$gene) + NoLegend()
```


## Percent expression of top markers across clusters

```{r, fig.width=16, fig.height=14}
top_5_markers = unique(unlist(tapply(cluster_markers$gene, cluster_markers$cluster, head, 5)))
DotPlot(scData.combined, features = top_5_markers , dot.scale = 4, cols = c("red", "blue"), assay = "RNA") +    RotatedAxis()
```

## Log2 Fold Changes of top 10 DE genes from each cluster

```{r, fig.width=12, fig.height=6}
plot_top_genes(scData.combined, cluster_markers, "Cluster markers")
```


```{r}

plot_genes <- function(dataset, genes) {
  
  all_genes =rownames(dataset@assays$RNA)
  
  for (g in genes) {
    if (!g %in% all_genes) {
      print(paste0("Gene ",g," has no data (possibly filtered out)."))
      next
    } else {
        p1  =FeaturePlot(object = dataset, features = g, reduction="umap", label = T)
        p2 = VlnPlot(dataset, features = g, ncol = 1, pt.size = 0.01)
        print(plot_grid(p1,p2, ncol=2,align="h"))
    }
  }
}
```


# Individual marker expression

## Global markers 

These are global cell markers for major cell lineages and types (B/T-cells, Mo/MF etc)

CD45 is Ptprc.  

We here add Cd3e/d/g as pan-T marker and Cd20 (Ms4a1) as mature B-cell marker.
Gata3, Cxcr6, Rora - T-cell markers


Endothelial - VWF, PECAM1  
Glial - S100B  
Pericytes - RGS5
Plasma cells - SDC1  
LEC - Lyve1
Fibroblasts - Col, Fn, Dcn

```{r, fig.width=14, fig.height=6}
# "Cd4","Cd3e","Cd3d","Cd3g", "Gata3","Cxcr6","Rora","Cd7",
sel_markers=(c("Ptprc","Epcam","Vim", "Vwf","Pecam1","S100b","Rgs5","Acta2", "Lyve1", "Col1a1","Col3a1","Fn","Dcn","Kit","Ano1"))
DefaultAssay(scData.combined) <- "RNA"

plot_genes(scData.combined, sel_markers)

```

## Stromal markers

S1-4 are from Kinchen et al human stroma data; MSC is from Jasso et al. 

S1: CCL8,FABP5  
S2: Sox6, VSTM2A,AGT,BMP2,POSTN  
S3: OGN, GSN, GREM1
S4: CCL9,NR2F2,IRF8  
MSC - Pdgfra, Smoc2

```{r, fig.width=14, fig.height=6}
sel_markers=human2mouse(c("CCL8","FABP5","SOX6","VSTM2A","AGT","BMP2","POSTN","OGN","GSN","GREM1","CCL9","NR2F2","IRF8","CCL19","CD74","PDGFRA","SMOC2"), db=homologeneData2)$mouseGene
DefaultAssay(scData.combined) <- "RNA"

plot_genes(scData.combined, sel_markers)
```

## MF markers

```{r, fig.width=14, fig.height=6}
sel_markers=human2mouse(c("ACTA2","TAGLN","ACTG2","MYH11"), db=homologeneData2)$mouseGene
DefaultAssay(scData.combined) <- "RNA"

plot_genes(scData.combined, sel_markers)
```

## Proliferation

```{r, fig.width=14, fig.height=6}
sel_markers=(c("Top2a","Mki67","Pcna"))
DefaultAssay(scData.combined) <- "RNA"

plot_genes(scData.combined, sel_markers)
```



## Mitochondrial transcript proportion

```{r}
FeaturePlot(object = scData.combined,  features = "percent_mito", reduction="umap", label=T)
```

```{r, fig.height=8}
tmp = data.frame(percent_mito = scData.combined$percent_mito, batch= scData.combined$orig.ident, group_label = scData.combined$condition )
ggplot(tmp, aes(fill=batch, x=group_label, y = percent_mito)) + geom_boxplot() + theme(axis.text.x = element_text(angle=90, hjust=1))

```


## Scores and RNA count per cluster

```{r, fig.width=10, fig.height=6}
VlnPlot(scData.combined, features = "percent_mito", ncol = 1, pt.size = 0.01)
#VlnPlot(scData.combined, features = "HALLMARK_APOPTOSIS", ncol = 1, pt.size = 0.01)
VlnPlot(scData.combined, features = "nCount_RNA", ncol = 1, pt.size = 0.01)
VlnPlot(scData.combined, features = "nFeature_RNA", ncol = 1, pt.size = 0.01)
```

```{r}
FeaturePlot(object = scData.combined,  features = "nFeature_RNA", reduction="umap", label=T)
```


# Final cluster assigment

```{r}
cluster_assignment = read.table("../../../Data/cluster_assignments/stroma_in_vivo/Jasso/Clusters_2023-01-10.txt", sep="\t",header=T, stringsAsFactors = F)
options(width = 180)
print(cluster_assignment)
rownames(cluster_assignment) = as.character(cluster_assignment$ClusterID)
scData.combined = AddMetaData(scData.combined, cluster_assignment[as.character(scData.combined$seurat_clusters),"Label"], col.name = "cluster_label")
scData.combined = AddMetaData(scData.combined, cluster_assignment[as.character(scData.combined$seurat_clusters),"Type"], col.name = "cluster_type")
```

```{r, fig.width=14, fig.height=8}
scData.combined = SetIdent(scData.combined, value="cluster_label" )
p = DimPlot(scData.combined, reduction = "umap", label = TRUE)
print(p)
```

## Distribution of clusters within samples


```{r}
m = table(scData.combined$cluster_label, scData.combined$sample_label)
m
```

```{r, fig.width=10,fig.height=7}
mm  = as.matrix.data.frame(m)
colnames(mm) = colnames(m)
rownames(mm) = rownames(m)

mm_r = sweep(mm, 2, apply(mm,2,sum), "/")

pheatmap(mm_r, cluster_cols = F, cluster_rows=F, main="Proportion of each cluster per sample")
```


## Distribution of clusters within conditions

```{r}
m = table(scData.combined$cluster_label, scData.combined$condition)
m
```

```{r}
mm  = as.matrix.data.frame(m)
colnames(mm) = colnames(m)
rownames(mm) = rownames(m)

mm_r = sweep(mm, 2, apply(mm,2,sum), "/")
options(width=200)
kable(round(mm_r,3))
```


```{r, fig.width=10,fig.height=7}
pheatmap(mm_r, cluster_cols = F, cluster_rows=F)
```

```{r, fig.width=14, fig.height=8}
mm_rel_ts = reshape2::melt(t(mm_r))
colnames(mm_rel_ts) = c("Group","Cluster","Proportion_cells")
mm_rel_ts$Cluster = factor(paste0("",mm_rel_ts$Cluster))

ggplot(mm_rel_ts, aes(x=Group, y = Proportion_cells, group=Cluster)) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle=60, hjust=1)) + facet_wrap(~ Cluster )

```

```{r}
save(scData.combined,file=file.path(data_storage_folder, "Colon_mesenchyme_mouse.combined_cluster_annotated.RData"))
```


# Export

```{r, eval=FALSE}
scData.combined <- SetIdent(scData.combined, value = "Celltype_Stroma_Atlas")
# if(rerun_analyses) {
  cluster_markers <- FindAllMarkers(object = scData.combined, only.pos = TRUE, assay = "RNA")
  save(cluster_markers, file=file.path(data_storage_folder, "All_markers_Celltype_Stroma_Atlas.Rdata"))
  write_xlsx(cluster_markers,file.path(result_folder, "All_markers_Celltype_Stroma_Atlas.xlsx"))
# } else {
#   load(file.path(data_storage_folder, "All_markers.Rdata"))
# }
```

```{r, eval=FALSE}
Stroma_simple = c("BEC 1"="BEC","BEC 2"="BEC","Fibroblast 1a"="Fib 1","Fibroblast 1b"="Fib 1","Fibroblast 1c"="Fib 1","Fibroblast 2a"="Fib 2","Fibroblast 2b"="Fib 2","Fibroblast 2c"="Fib 2","Fibroblast 3a"="Fib 3","Fibroblast 3b"="Fib 3","Glial"="Glial","ICC 1"="ICC","ICC 2"="ICC","LEC"="LEC","MSC"="MSC","Pericytes"="Pericytes","SMC 1"="SMC","SMC 2"="SMC")
scData.combined$celltype_stroma_simple <- Stroma_simple[as.character(scData.combined$Celltype_Stroma_Atlas)]

DimPlot(scData.combined, group.by="celltype_stroma_simple", label=T)

scData.combined <- SetIdent(scData.combined, value = "celltype_stroma_simple")
cluster_markers <- FindAllMarkers(object = scData.combined, only.pos = TRUE, assay = "RNA")
save(cluster_markers, file=file.path(data_storage_folder, "All_markers_Celltype_Stroma_Atlas_simple.Rdata"))
write_xlsx(cluster_markers,file.path(result_folder, "All_markers_Celltype_Stroma_Atlas_simple.xlsx"))

```

# Software versions

```{r}
sessionInfo()
```














