---
title: "Cluster identification - Assembloids clusters, mouse"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---


```{r, message=FALSE}
rm(list=ls())
library(Seurat)
library(ggplot2)

library(readxl)
library(writexl)
library(knitr)
library(data.table)
library(cowplot)
library(magrittr)
library(dplyr)

library(homologene)

library(pheatmap)

library(grid)

# library(clusterProfiler)
# 
# options(connectionObserver = NULL)
# library(org.Hs.eg.db)
# library(ReactomePA)
# library(msigdbr)

data_storage_folder = "../../Results/stromal/"
result_folder = data_storage_folder

library(future)
options("future.globals.maxSize"=2**33)
plan("multisession", workers = 8)
plan("sequential")

options("save.defaults" = list(compress=TRUE))

#source("DotPlot_fixed.R")


#ref_folder = "../../References/"

rerun_analyses = F
rerun_gsva = F
```


```{r}
scData.combined <- readRDS(file.path(data_storage_folder, "Assembloids_combined_SCT_CC_adj.Rdata"))
set.seed(1234)
DefaultAssay(scData.combined) <- "integrated"
scData.combined <- FindClusters(scData.combined, resolution = 0.4)  # modify resolution to change the number of clusters
DefaultAssay(scData.combined) <- "RNA"


```


<!-- ```{r} -->
<!-- in_vivo_combined_meta = read.table("Metadata_Jasso_Kinchen_Assembloids_SCT_Harmony.txt", sep="\t", header=T) -->
<!-- rownames(in_vivo_combined_meta) <- in_vivo_combined_meta$cell_id -->

<!-- scData.combined$clusters_with_Jasso <- in_vivo_combined_meta[Cells(scData.combined), "cluster_label"] -->
<!-- ``` -->


# Cluster overview

# New clusters

```{r, fig.width=10, fig.height=6}
DimPlot(scData.combined, reduction = "umap", label = TRUE, group.by = "seurat_clusters")
```

<!-- ## Assembloids only -->

<!-- ```{r, fig.width=10, fig.height=6} -->
<!-- scData.combined <- SetIdent(scData.combined, value="cluster_label") -->
<!-- DimPlot(scData.combined, reduction = "umap", label = TRUE) -->
<!-- ``` -->

<!-- ```{r, fig.width=10, fig.height=6} -->
<!-- scData.combined <- SetIdent(scData.combined, value="clusters_with_Jasso") -->
<!-- DimPlot(scData.combined, reduction = "umap", label = TRUE) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- table(scData.combined$cluster_label, scData.combined$clusters_with_Jasso) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- tt = table(scData.combined$seurat_clusters, scData.combined$clusters_with_Jasso) -->
<!-- tt -->
<!-- ``` -->

<!-- ```{r} -->
<!-- pheatmap(tt) -->
<!-- ``` -->



# Final cluster assigment

```{r}
cluster_assignment = read.table("../../Data/cluster_assignments/stromal/Clusters_Assembloids_SCT_CCadj_2023-02-09_simple_clusters.txt", sep="\t",header=T, stringsAsFactors = F)
options(width = 180)
print(cluster_assignment)
rownames(cluster_assignment) = as.character(cluster_assignment$ClusterID)
scData.combined = AddMetaData(scData.combined, cluster_assignment[as.character(scData.combined$seurat_clusters),"Label"], col.name = "cluster_label")
scData.combined = AddMetaData(scData.combined, cluster_assignment[as.character(scData.combined$seurat_clusters),"Type"], col.name = "cluster_type")

scData.combined$cluster_label = scData.combined$cluster_label |> (\(x) factor(x, levels = sort(unique(x))))()
```

```{r, fig.width=8, fig.height=6}
scData.combined = SetIdent(scData.combined, value="cluster_label" )
p = DimPlot(scData.combined, reduction = "umap", label = TRUE)
print(p)
```


```{r}
saveRDS(scData.combined, file=file.path(data_storage_folder, "Assembloids_combined_SCT_CC_adj_annotated.rds"))
```


# Marker genes for clusters

```{r}
scData.combined <- SetIdent(scData.combined, value="cluster_label")
if(rerun_analyses) {
  cluster_markers <- FindAllMarkers(object = scData.combined, only.pos = TRUE, assay = "RNA")
  save(cluster_markers, file=file.path(data_storage_folder, "All_markers_assembloids_V2_simple_clusters.Rdata"))
  write_xlsx(cluster_markers,file.path(result_folder, "All_markers_assembloids_V2_simple_clusters.xlsx"))
} else {
  load(file.path(data_storage_folder, "All_markers_assembloids_V2_simple_clusters.Rdata"))
}
```

```{r}
plot_top_genes <- function(data, dge_results_all, title) {
  for (n in sort(unique(dge_results_all$cluster))) {
    tmp = subset(dge_results_all, cluster==n)
    if(nrow(tmp)<1) next
    tmp_up = subset(tmp, avg_log2FC > 0)
    if(nrow(tmp_up)<1) next
    tmp_up$direction = "up"
    #tmp_down = subset(tmp, avg_log2FC < 0)
    #tmp_down$direction = "down"

    #final_top_genes = rbind(head(tmp_up,10), head(tmp_down,10))
    final_top_genes = head(tmp_up, 10)
    final_top_genes = final_top_genes[order(final_top_genes$avg_log2FC),]
    final_top_genes$gene = factor(final_top_genes$gene, levels = final_top_genes$gene)
    p = ggplot(final_top_genes, aes(y=avg_log2FC, x=gene, fill=direction)) + geom_bar(stat="identity") + ggtitle(paste0(n, ", ", title)) + coord_flip() + theme(axis.text.y =  element_text(size=12) )
    print(p)

  }
}
```


```{r}
cluster_markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> top10
```


## overview (10 top genes for each cluster)

```{r, fig.width=14, fig.height=12}
DoHeatmap(scData.combined, features = top10$gene) + NoLegend()
```


## Percent expression of top markers across clusters

```{r, fig.width=16, fig.height=14}
top_5_markers = unique(unlist(tapply(cluster_markers$gene, cluster_markers$cluster, head, 5)))
DotPlot(scData.combined, features = top_5_markers , dot.scale = 4, cols = c("red", "blue"), assay = "RNA") +    RotatedAxis()
```

## Log2 Fold Changes of top 10 DE genes from each cluster

```{r, fig.width=12, fig.height=6}
plot_top_genes(scData.combined, cluster_markers, "Cluster markers")
```


```{r}

plot_genes <- function(dataset, genes) {
  
  all_genes =rownames(dataset@assays$RNA)
  
  for (g in genes) {
    if (!g %in% all_genes) {
      print(paste0("Gene ",g," has no data (possibly filtered out)."))
      next
    } else {
        p1  =FeaturePlot(object = dataset, features = g, reduction="umap", label = T, order=T)
        p2 = VlnPlot(dataset, features = g, ncol = 1, pt.size = 0.01)
        print(plot_grid(p1,p2, ncol=2,align="h"))
    }
  }
}
```

```{r}
#scData.combined <- SetIdent(scData.combined, value="seurat_clusters")
```


# Individual marker expression

## Global markers 

These are global cell markers for major cell lineages and types (B/T-cells, Mo/MF etc)

CD45 is Ptprc.  

We here add Cd3e/d/g as pan-T marker and Cd20 (Ms4a1) as mature B-cell marker.
Gata3, Cxcr6, Rora - T-cell markers


Endothelial - VWF, PECAM1  
Glial - S100B  
Pericytes - RGS5  
Plasma cells - SDC1  
ICC - Kit, Ano1 (Jasso et al. )


```{r, fig.width=14, fig.height=6}

sel_markers=(c("Ptprc","Cd4","Cd3e","Cd3d","Cd3g", "Gata3","Cxcr6","Rora","Cd7","Cd19","Ms4a1","Epcam","Vim", "Vwf","Pecam1","S100b","Rgs5","Sdc1", "Hcar2", "Kit","Ano1"))
DefaultAssay(scData.combined) <- "RNA"

plot_genes(scData.combined, sel_markers)

```

```{r, fig.width=15, fig.height=3}

DotPlot(scData.combined, features = sel_markers, scale =T, cols = "RdYlGn", assay = "RNA")

```

## Stromal markers


### Subpopulation markers from McCarthy et al. (Nat Cell Biol 2020, p.1030 ff)

```{r, fig.width=14, fig.height=6}
sel_markers = unlist(strsplit("Wnt2b, Wnt4, Wnt5a, Rspo1, Rspo2, Rspo3, Dkk2, Dkk3, Sfrp1, Frzb, Wif1, Grem1, Grem2, Chrd, Bmp2, Bmp4, Bmp5, Bmp7, Cd34, Pdgfra, Foxl1, Gli1, Des, Id1, Id2", ", "))
#sel_markers=human2mouse(sel_genes, db=homologeneData2)$mouseGene
DefaultAssay(scData.combined) <- "RNA"

plot_genes(scData.combined, sel_markers)
```

```{r, fig.width=15, fig.height=3}

DotPlot(scData.combined, features = sel_markers, scale =T, cols = "RdYlGn", assay = "RNA")

```


### Subpopulation markers from Kinchen et al. (Fig. 3)

Str 1: Cxcl12, Adamdec1, Sfrp2, Hcar2, Nxf7, 
Str 2: Wnt5a, Bmp2, Sox6, Fgf9
Str 3: Maff, Cd55, Pcolce12, Ackr4

```{r, fig.width=14, fig.height=6}
sel_markers = c("Cxcl12", "Adamdec1","Pcolce2","Ackr4", "Wnt5a","Bmp2","Sox6", "Fgf9", "Maff","Cd55", "Hcar2", "Nxf7", "Sfrp2")
#sel_markers=human2mouse(sel_genes, db=homologeneData2)$mouseGene
DefaultAssay(scData.combined) <- "RNA"

plot_genes(scData.combined, sel_markers)
```

```{r, fig.width=12, fig.height=3}

DotPlot(scData.combined, features = sel_markers, scale =T, cols = "RdYlGn", assay = "RNA")

```

### Subpopulation markers from Jasso et al. 

Str 1: 
Str 2: Procr, Sox6
Str 3: 

```{r, fig.width=14, fig.height=6}
sel_markers = c("Procr","Sox6")
#sel_markers=human2mouse(sel_genes, db=homologeneData2)$mouseGene
DefaultAssay(scData.combined) <- "RNA"

plot_genes(scData.combined, sel_markers)
```

```{r, fig.width=12, fig.height=3}

DotPlot(scData.combined, features = sel_markers, scale =T, cols = "RdYlGn", assay = "RNA")

```


## MF markers

```{r, fig.width=14, fig.height=6}
sel_markers=human2mouse(c("ACTA2","TAGLN","ACTG2","MYH11"), db=homologeneData2)$mouseGene
DefaultAssay(scData.combined) <- "RNA"

plot_genes(scData.combined, sel_markers)
```

```{r, fig.width=12, fig.height=3}

DotPlot(scData.combined, features = sel_markers, scale =T, cols = "RdYlGn", assay = "RNA")

```

## Proliferation

```{r, fig.width=14, fig.height=6}
sel_markers=(c("Top2a","Mki67","Pcna"))
DefaultAssay(scData.combined) <- "RNA"

plot_genes(scData.combined, sel_markers)
```

```{r, fig.width=12, fig.height=3}

DotPlot(scData.combined, features = sel_markers, scale =T, cols = "RdYlGn", assay = "RNA")

```


# Export

Cluster assignations have been exported to file *Metadata_Assembloids_SCT_CCadj_simple_clusters.txt*.

```{r}
tmp = scData.combined@meta.data
tmp$cell_id = rownames(tmp)
write.table( tmp , file=file.path(data_storage_folder, "Metadata_Assembloids_SCT_CCadj_simple_clusters.txt"), sep="\t", row.names = F, quote = F)
```


# Jasso et al signatures

# Simplified original Jasso et al. stroma clusters

```{r}
tmp_env = new.env()
load("../../Results/Jasso_Stroma/All_markers_Celltype_Stroma_Atlas_simple.Rdata", envir = tmp_env)
tmp_markers = get("cluster_markers", tmp_env)
tmp_markers = subset(tmp_markers, p_val_adj < 0.05 & avg_log2FC > 0.5)
tmp_gs = split(tmp_markers$gene, tmp_markers$cluster)
names(tmp_gs) = paste0("Jasso_", names(tmp_gs))
```


```{r}
library(GSVA)
rerun=rerun_gsva
gsva_score_file = file.path(data_storage_folder, file="Assembloids_SCT_CCadj_Jasso_Stroma_Orig_Simple_signatures_GSVA.rds")
if(rerun) {
  scores = gsva(scData.combined@assays$SCT@data, gset.idx.list = tmp_gs, parallel.sz = 12)
  saveRDS(scores, file=gsva_score_file)
} else {
  scores = readRDS(gsva_score_file)
}

for(ss in rownames(scores)) {
  scData.combined@meta.data[[ss]] = scores[ss, Cells(scData.combined)]
}

```

```{r}
for(ss in rownames(scores)) print(FeaturePlot(scData.combined, features = ss, order=T, label=T))
```

```{r}
for(ss in sort(rownames(scores))) print(VlnPlot(scData.combined, features = ss, pt.size = 0) + geom_hline(yintercept = 0))
```



```{r, fig.height=4, fig.width=8}
DotPlot(scData.combined, features = sort(rownames(scores)), cols = "RdYlBu") + theme(axis.text.x = element_text(angle=45, hjust=1))
DotPlot(scData.combined, features = sort(rownames(scores))) + theme(axis.text.x = element_text(angle=45, hjust=1))
```

# Kinchen + Jasso combined clusters

```{r}
tmp_env = new.env()
load("../../Results/Jasso_Kinchen_combined/All_markers_Kinchen_Jasso.Rdata", envir = tmp_env)
tmp_markers = get("cluster_markers", tmp_env)
tmp_markers = subset(tmp_markers, p_val_adj < 0.05 & avg_log2FC > 0.5)
tmp_gs = split(tmp_markers$gene, tmp_markers$cluster)
names(tmp_gs) = paste0("KinchenJasso_", names(tmp_gs))
```

```{r}
library(GSVA)

rerun=rerun_gsva
gsva_score_file = file.path(data_storage_folder, file="Assembloids_SCT_CCadj_KinchenJasso_Stroma_signatures_GSVA.rds")
if(rerun) {
  scores = gsva(scData.combined@assays$SCT@data, gset.idx.list = tmp_gs, parallel.sz = 12)
  saveRDS(scores, file=gsva_score_file)
} else {
  scores = readRDS(gsva_score_file)
}

for(ss in rownames(scores)) {
  scData.combined@meta.data[[ss]] = scores[ss, Cells(scData.combined)]
}

```

```{r}
for(ss in sort(rownames(scores))) print(FeaturePlot(scData.combined, features = ss, order=T, label=T))
```

```{r}
for(ss in sort(rownames(scores))) print(VlnPlot(scData.combined, features = ss, pt.size = 0) + geom_hline(yintercept = 0))
```



```{r, fig.height=5, fig.width=8}
DotPlot(scData.combined, features = sort(rownames(scores)), cols = "RdYlBu") + theme(axis.text.x = element_text(angle=45, hjust=1))
DotPlot(scData.combined, features = sort(rownames(scores))) + theme(axis.text.x = element_text(angle=45, hjust=1))
```


# Software versions

```{r}
sessionInfo()
```














